Dish: Distribution-oblivious Shell Scripting

### 2019-11-22 pldi/draft.tex
Pass on 4

In order to effectively apply optimizations and distribute a shell
script, it first has to be translated to a manipulable
representation. As mentioned in \ref{}, the two main abstractions of
the shell, are files that contain data, and commands that communicate
through these files. We propose a slight variation of the standard
dataflow graph model, that is commonly found in stream processing
systems, as the representation of a distributed shell computation. The
main benefit of the dataflow graph model is that it clearly exposes
task parallelism, as different nodes can be independent processing
workers. In our case each node represents a command, and each edge
represents a file.

@@

Nodes $f$ of the graph represent functions from one (possibly empty)
input stream to an output stream $f : D^* \rightarrow D^*$. $D$
represents the basic data type of a line of characters. This
representation captures the majority of shell commands. The nodes are
monotonic, meaning that they cannot retract their output. More
formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there exists
$y'$ such that $f(x.x') = y.y'$, where $.$ is standard concatenation
and will be omitted when obvious. An example of a node is the
following command \kk{Does it need ref? Also maybe this example is too
  trivial and can be removed.}:

@@

Edges in the dataflow graph represent files, the basic data
abstraction of the shell. They are used as communication channels
between nodes in the graph, and as the input or output of the whole
graph. Edges are represented as possibly unbounded streams of type
$D*$. They could either refer to a specific named file in the file
system, or just be FIFO pipes that used for interprocess communication
\TODO{Are there any other file types, such as URLs, ...?}.

@@

\TODO{Add an example of a command that has two outputs and one goes to
  one command, while another output goes to a different one. Maybe
  only show the graph here and not the command itself.}

@@

In contrast, our dataflow model inherently supports the data
parallelism that is found in shell programs. More precisely, data
parallelism can be easily exposed as task parallelism in our model. A
very important consequence of this is that we can iteratively apply
optimizations that expose data parallelism, or even revert them when
the resulting dataflow graph is too wide. This flexibility enables
unified reasoning and implementation of the unoptimized and optimized
version of the dataflow graph. \kk{This is exciting, and maybe it
  should be moved up (or rephrased in the introduction)}

Before describing the different types of parallelization
optimizations, we formalize our intuition about stateless and pure
commands.
\nv{TODOs
  (i) highlight earlier the antithesis between the formalisms in this section and the informalities of the previous (me);
  (ii) try to add a proof \emph{sketch} pointing to these formalisms (to show their value), and we can point to the TR for the longer proof.
}
 Recall the example \ref{} from section Categories
\ref{}. This example shows the stateless command x being applied to
...\kk{Complete this after we have an example}. Formally, a commmand
$f$ is stateless if it is ``linear'' with respect to concatenation,
i.e. the following equation holds:

@@

Similarly, every pure command $f$ that can be written as a map $m$ and a
reduce $r$ step satisfies the following:

>>>>

In order to effectively apply optimizations and distribute a shell
script, it first has to be translated to a manipulable
representation. As mentioned in \Cref{bg:pipelines}, the two main
abstractions of the shell, are files that contain data, and commands
that communicate through these files. We propose a slight variation of
the standard dataflow graph model, that is commonly found in stream
processing systems, as the representation of a distributed shell
computation. The main benefit of the dataflow graph model is that it
clearly exposes task parallelism, as different nodes can be
independent processing workers. In our case each node represents a
command, and each edge represents a file.

@@

Nodes $f$ of the graph represent functions from one (possibly empty)
input stream to an output stream $f : D^* \rightarrow D^*$. $D$
represents the basic data type of a line of characters. This
representation captures the majority of shell commands. The nodes are
monotonic, meaning that they cannot retract their output. More
formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there exists
$y'$ such that $f(x.x') = y.y'$, where $.$ is standard concatenation
and will be omitted when obvious. An example of a node is the
following command:

@@

Edges in the dataflow graph represent files, the basic data
abstraction of the shell. They are used as communication channels
between nodes in the graph, and as the input or output of the whole
graph. Edges are represented as possibly unbounded streams of type
$D*$. They could either refer to a specific named file in the file
system, or just be FIFO pipes that used for interprocess communication
%% \TODO{Are there any other file types, such as URLs, ...?}.

@@

\TODO{Add an example of a command that has two outputs, one of which
  goes to one command, while another goes to a different
  command. Maybe only show the graph here and not the command itself.}

@@

In contrast, our dataflow model inherently supports the data
parallelism that is found in shell programs. More precisely, data
parallelism can be easily exposed as task parallelism in our model. A
very important consequence of this is that we can iteratively apply
optimizations that expose data parallelism, or even revert them when
the resulting dataflow graph is too wide. This flexibility enables
unified reasoning and implementation of the unoptimized and optimized
version of the dataflow graph. \kk{This is interesting, and should be
  moved up (or rephrased in the introduction)}

Before describing the different types of parallelization
optimizations, we formalize the intuition about stateless and pure
commands that was described in \Cref{cmd}.
\nv{TODOs
  (i) highlight earlier the antithesis between the formalisms in this section and the informalities of the previous (me);
  (ii) try to add a proof \emph{sketch} pointing to these formalisms (to show their value), and we can point to the TR for the longer proof.
}
%
As mentioned there, stateless commands operate independently on
individual elements of the stream (characters, lines, or files)
without maintaining any state. Some examples include \ttt{grep},
\ttt{basename}, and \ttt{tr}. Formally, a commmand $f$ is stateless if
it is ``linear'' with respect to concatenation, i.e. the following
equation holds:

@@

Similarly, it was mentioned that some pure commands (such as
\ttt{sort} and \ttt{wc}) can be parallelized using divide-and-conquer
techniques. More formally, there pure commands $f$ can be implemented
as a map $m$ and a reduce $r$ function, satisfying the following:
